---
title: "HIV/AIDS Surveillance Data Analysis"
author: "Xuan Lu and Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    css: style.css
---

## Project Proposal Summary

### The Team
- Junjie Hu (jh4617)
- Ruixi Li (rl33298)
- Xuan Lu (xl3214)
- Tianyun Jiang (tj2519)
- Qinting Shen (qs2261)

### Tentative Project Title
HIV/AIDS Surveillance Data Analysis across Countries

### Motivation
The primary goal is to analyze information on HIV prevalence, incidence rates, and AIDS mortality from available studies, identifying the most vulnerable gender, sub-population, and geographic region.

### Intended Final Products
Multiple dashboards/web pages reflecting the prevalence and incidence rates of HIV across countries and subgroups.

### Anticipated Data Sources
United States Census’s HIV/AIDS Surveillance Data Base

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Data Cleaning and Preparation
```{r clean data}
# Load the dataset (assuming you have already read it into 'hiv_df')
hiv_df <- read.csv("hiv_joined_data.csv") |>
  janitor::clean_names()

# View the first few rows of the dataset
head(hiv_df)

# Display the column names
colnames(hiv_df)

# Filter data for new HIV incidences and AIDS mortality
hiv_df_filtered <- hiv_df |>
    dplyr::select(new_hiv_incidences, aids_mortality, age, gender, sub_population, geographic_region) |>
    dplyr::filter(!is.na(new_hiv_incidences) & !is.na(aids_mortality))

# Further filter by demographic groups if necessary
```

## Exploratory Analysis
```{r EDA}
# Histogram for new HIV incidences
ggplot(hiv_df_filtered, aes(x = new_hiv_incidences)) +
    geom_histogram(binwidth = 1, fill = "blue", color = "black")

# Summary statistics for AIDS mortality
summary(hiv_df_filtered$aids_mortality)
```

## Statistical Testing
```{r Stat Tests}
# Load necessary libraries
library(broom)  # For tidy statistical summaries

# Chi-square test for categorical variables (e.g., gender)
chi_sq_result <- chisq.test(hiv_df_filtered$gender, hiv_df_filtered$new_hiv_incidences)
print(glue::glue("Chi-Square Test for Gender and New HIV Incidences:\n"))
print(tidy(chi_sq_result))

# ANOVA for numeric variables (e.g., age)
aov_result <- aov(aids_mortality ~ age, data = hiv_df_filtered)
print(glue::glue("\nANOVA for AIDS Mortality by Age:\n"))
print(tidy(aov_result))

# T-test (e.g., comparing two age groups)
t_test_result <- t.test(aids_mortality ~ age_group, data = hiv_df_filtered)
print(glue::glue("\nT-Test for AIDS Mortality by Age Group:\n"))
print(tidy(t_test_result))

# Regression (e.g., multiple variable regression)
lm_result <- lm(aids_mortality ~ age + gender + geographic_region, data = hiv_df_filtered)
print(glue::glue("\nLinear Regression for AIDS Mortality:\n"))
print(tidy(lm_result))
```

- Chi-Square Test: Used for testing relationships between categorical variables.
- ANOVA: Analyzes differences between group means and their associated procedures.
- T-Test: Compares the means of two groups.
- Linear Regression: Models the relationship between a dependent variable and one or more independent variables.

### Validation of Data:

- Ensure hiv_df_filtered contains the necessary columns (gender, new_hiv_incidences, aids_mortality, age, age_group, geographic_region).
- Validate the assumptions for each statistical test (e.g., normality for t-test and ANOVA, independence for Chi-square).

## Visualization
```{r visualization}
library(plotly)
library(ggplot2)

# HIV Incidence Rates by Country - Scattergeo Map
map_plot <- plot_ly(data = hiv_df_filtered, lat = ~latitude, lon = ~longitude, text = ~country, type = "scattergeo") |>
            layout(title = 'HIV Incidence Rates by Country',
                   geo = list(projection = list(type = 'natural earth')))

# HIV Incidence Rates Over Time by Subpopulation - Line Chart
time_plot <- plot_ly(hiv_df_filtered, x = ~year, y = ~new_hiv_incidences, color = ~sub_population, type = 'scatter', mode = 'lines') |>
             layout(title = 'HIV Incidence Rates Over Time by Subpopulation')

# Difference in HIV Incidence Rates Across Gender - Bar Chart
gender_plot <- ggplot(hiv_df_filtered, aes(x = gender, y = new_hiv_incidences)) +
               geom_bar(stat = "identity", fill = "steelblue") +
               theme_minimal() +
               labs(title = 'Difference in HIV Incidence Rates Across Gender')

# HIV Incidence Rates in Capital City vs Rural Area - Bar Chart
location_plot <- ggplot(hiv_df_filtered, aes(x = location_type, y = new_hiv_incidences)) +
                 geom_bar(stat = "identity", fill = "coral") +
                 theme_minimal() +
                 labs(title = 'HIV Incidence Rates in Capital City vs Rural Area')

# HIV Incidence Rates Across Continents - Bar Chart
continent_plot <- ggplot(hiv_df_filtered, aes(x = continent, y = new_hiv_incidences)) +
                  geom_bar(stat = "identity", fill = "lightgreen") +
                  theme_minimal() +
                  labs(title = 'HIV Incidence Rates Across Continents')

# Print plots (optional, based on your setup)
# print(map_plot)
# print(time_plot)
# print(gender_plot)
# print(location_plot)
# print(continent_plot)
```

## Tables
```{r Tables}
# Load necessary libraries
library(dplyr)
library(knitr)

# Table for countries/regions with highest HIV prevalences and incidence rates
top_countries <- hiv_df_filtered |>
    group_by(country) |>
    summarise(Average_Incidence = mean(new_hiv_incidences, na.rm = TRUE)) |>
    arrange(desc(Average_Incidence)) |>
    head(n = 10)  # Get top 10 countries

# Display the table with kable for a professional look
kable(top_countries, caption = "Top 10 Countries/Regions with Highest HIV Incidence Rates", 
      format = "html", table.attr = "style='width:100%;'", 
      col.names = c("Country/Region", "Average Incidence Rate"))
```

