---
title: "HIV/AIDS Surveillance Data Analysis"
author: "Xuan Lu and Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    css: style.css
editor_options: 
  chunk_output_type: console
---

## Project Proposal Summary

### The Team
- Junjie Hu (jh4617)
- Ruixi Li (rl3328)
- Xuan Lu (xl3214)
- Tianyun Jiang (tj2519)
- Qinting Shen (qs2261)

### Tentative Project Title
HIV/AIDS Surveillance Data Analysis across Countries

### Motivation
The primary goal is to analyze information on HIV prevalence, incidence rates, and AIDS mortality from available studies, identifying the most vulnerable gender, sub-population, and geographic region.

### Intended Final Products
Multiple dashboards/web pages reflecting the prevalence and incidence rates of HIV across countries and subgroups.

### Anticipated Data Sources
United States Census’s HIV/AIDS Surveillance Data Base

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```




## Data Cleaning and Preparation

```{r clean data}
# Incidence

incidence <- read.csv("surveillance/data/hiv_incidence.csv") |>
  janitor::clean_names() |>
  select(-c(no_cases,no_deaths,prev_rate)) |>
  rename(subpop = population_subgroup) |>
  mutate(geographic_area = ifelse(
    str_detect(geographic_area,"rural"),"rural", 
    ifelse(
      str_detect(geographic_area,"semi"),"semiurban",
    ifelse(
      str_detect(geographic_area,"urban"),"urban", geographic_area
    ))))

## recode subpopulation

# sexually active, TB, STI : other non-representative
# pregnant women: other patients
# sex worker&clients: ?
# Patients:?
# HIV+ individuals:?
incidence = incidence |>
  mutate(
    subpop_pooled = ifelse(
     str_detect(subpop, "(?i)police|military"), "Military/Armed Forces",
     ifelse(
       str_detect(subpop, "(?i)child|pediatric"), "Children",
     ifelse(
       str_detect(subpop, "(?i)drug|IVDU") & !str_detect(subpop, "(?i)STI|homo|prisoner|partner|sex worker"),  "Intravenous Drug Users/Needle Sharers",
     ifelse(
       str_detect(subpop, "(?i)sex worker|bar") & !str_detect(subpop,"(?i)trans|IVDU|homo|MSM|drug|client"),"Sex workers",
     ifelse(
       str_detect(subpop, "(?i)transgender|homo|MSM|gay|TB|STI|sexually|prisoner|infants born|HIV2+|high risk|of HIV") & !str_detect(subpop, "(?i)sex worker|IVDU|drug|Testing center attendees"), "Other Non-Representative",
      ifelse(
       str_detect(subpop, "(?i)pts.|pregnant|Mothers") & !str_detect(subpop, "(?i)sex worker|IVDU|STI|homo|TB"), "Other Patients",
      ifelse(
       str_detect(subpop, "(?i)Blood|port|Testing center attendees|employer|contractor|textile|Wives|HIV-") , "General Population","Two Known Mixed Groups"))))))))

# unique(incidence$subpop[str_detect(incidence$subpop,"patient")]) for categorization
# unique(incidence$subpop[incidence$subpop_pooled=="Two Known Mixed Groups"])
incidence = incidence |>
  mutate(subpop_pooled = ifelse(
    subpop %in% c("Bisexuals","Clients of sex workers","High risk individuals","Partners of drug users","Blood transfusion recipients"),"Other Non-Representative",subpop_pooled),
    subpop_pooled = ifelse(
      subpop == "Sex workers & clients","Sex workers",subpop_pooled),
    subpop_pooled = ifelse(
      subpop == "Patients","Other Patients",subpop_pooled),
    subpop_pooled = ifelse(
      subpop %in% c("Heterosexuals","Adults","Individuals","Women","Various groups","Residents","Unspecified population","Truck drivers","Others","Adults - circumcised","Adults - uncircumcised","Factory workers", "General population", "Low risk groups","Fishermen","Workers","Esquineros","Trucking companies employees","Youths","Hospitality girls","Adolescents","Volunteers","Controls","Rural population","Employees","Commercial bank employees"),"General Population",subpop_pooled))
incidence |>filter(subpop == "Patients")
incidence |> filter(str_detect(title,"Estimation of HIV-1 Incidence with BED-CIAE among Clinical Patients in Liangshan Yi Autonomous Prefecture"))

check=incidence |>
  filter(subpop_pooled != "Two Known Mixed Groups") |>
  group_by(subpop_pooled,subpop) |>
  summarise(n=n())
  
incidence|>
  filter(subpop_pooled == "Two Known Mixed Groups") 
     

# prevalence

prevalence <- read.csv("surveillance/data/hiv_prevalence.csv") |>
  janitor::clean_names() |>
  select(-c(no_cases,no_deaths,inc_rate))|>
  rename(subpop = population_subgroup)
```

## Exploratory Analysis
```{r EDA}
# Histogram for HIV incidences

incidence |>
  ggplot(aes(x = inc_rate)) + geom_histogram(binwidth = 1, fill = "blue", color = "black")

# add a small constant 0.01 to each value and conduct log transformation

log_inc = log(pull(incidence,inc_rate)+0.01)

incidence |>
  ggplot(aes(x = log_inc)) + geom_histogram(binwidth = 1, fill = "blue", color = "black")

# Histogram for HIV prevalences

POR = log(pull(prevalence,prev_rate)/(1-pull(prevalence,prev_rate)))
prevalence |>
  ggplot(aes(x = prev_rate)) + geom_histogram(binwidth = 1, fill = "blue", color = "black")

# add a small constant 0.01 to each value and conduct log transformation

log_prev = log(pull(prevalence,prev_rate)+0.01)

prevalence |>
  ggplot(aes(x = log_prev)) + geom_histogram(binwidth = 1, fill = "blue", color = "black")


tri = (pull(prevalence,prev_rate))^(1/2)





prevalence |>
  ggplot(aes(x = tri)) + geom_histogram(binwidth = 1, fill = "blue", color = "black")


```

## Statistical Testing
```{r Stat Tests}
# Load necessary libraries
library(broom)  # For tidy statistical summaries

# ANOVA for numeric variables (e.g., age)
aov_prev = aov(prev_rate ~ sex, data = prevalence) |> broom::tidy()
aov_inc = aov(inc_rate ~ sex, data = incidence) |> broom::tidy()
print(glue::glue("\nANOVA for HIV Prevalence Rate by Age:\n"))
print(aov_prev)
print(glue::glue("\nANOVA for HIV Incidence Rate by Age:\n"))
print(aov_inc)


# t test: Difference in HIV incidence rate between the circumcised and uncircumcised 

circumcise = 
  incidence |>
  mutate(
    subpop = recode(subpop,
                                 'Adults - circumcised' = 'circumcised',
                                 'Adults - uncircumcised' = 'uncircumcised')) |>
  filter(subpop %in% c('circumcised','uncircumcised'))

ttest_result = t.test(inc_rate ~ subpop, data = circumcise) |> broom::tidy()
print(glue::glue("\nT-Test for HIV incidence by circumcision status among adults:\n"))
print(ttest_result)

```
According to the results of ttest, we have sufficient evidence to conclude that there's difference in incidence rate between the circumcised and uncircumcised adults. the uncircumcised adults are more likely to get HIV compared with the circumcised adults.





- Chi-Square Test: Used for testing relationships between categorical variables.
- ANOVA: Analyzes differences between group means and their associated procedures.
- T-Test: Compares the means of two groups.
- Linear Regression: Models the relationship between a dependent variable and one or more independent variables.

### Validation of Data:

- Ensure hiv_df_filtered contains the necessary columns (gender, new_hiv_incidences, aids_mortality, age, age_group, geographic_region).
- Validate the assumptions for each statistical test (e.g., normality for t-test and ANOVA, independence for Chi-square).

## Visualization
```{r visualization}
library(plotly)
library(ggplot2)

# A map reports the HIV prevalence and incidence rates in each country
map_inc = incidence |>
  plot_ly(type = 'choropleth', locations = ~country_code,
               z = ~inc_rate, text = ~paste(country, ': ', inc_rate, '%'),
               hoverinfo = 'text',color = ~inc_rate, colors = "Reds") |>
  layout(title = 'HIV Incidence Rates by country',
         geo = list(projection = list(type = 'orthographic')))

map_inc

map_prev = prevalence |>
  plot_ly(type = 'choropleth', locations = ~country_code,
               z = ~prev_rate, text = ~paste(country, ': ', prev_rate, '%'),
               hoverinfo = 'text',color = ~prev_rate, colors = "Reds") |>
  layout(title = 'HIV Prevalence Rates by country',
         geo = list(projection = list(type = 'orthographic')))
map_prev


# HIV Incidence Rates by circumcision status among adults

circumcise_plot = circumcise |>
  plot_ly(x = ~subpop, y = ~inc_rate, type = 'box') |>
  layout(title = 'HIV Incidence Rates by circumcision status among adults',
         yaxis = list(title = 'Incidence Rate(%)'),
         xaxis = list(title = 'Circumcision Status'))

circumcise_plot

# Difference in HIV Prevalence and Incidence Rates Across Gender
prev_sex = prevalence |>
  group_by(sex) |>
  summarize(median=median(prev_rate), n = n()) |>
  mutate(text_label = str_c("Sex:", sex, " \nMedian:", median, " Sample Size:", n))

sex_plot = incidence |>
  group_by(sex) |>
  summarise(median = median(inc_rate), n = n()) |>
  mutate(text_label = str_c("Sex:", sex, " \nMedian:", median, " Sample Size:", n)) |>
  plot_ly(x = ~sex, y = ~median, size = ~n, type = 'scatter', mode = 'markers', name = 'Incidence Rate',text = ~text_label, hoverinfo = 'text') |>
  add_trace(data = prev_sex, x = ~sex, y = ~median, size = ~n, type = 'scatter', name = 'Prevalence Rate',text = ~text_label, hoverinfo = 'text') |>
  layout(title = 'Median Incidence and Prevalence Rates by Sex',
         xaxis = list(title = 'Sex'),
         yaxis = list(title = 'Median Rate'))

print(sex_plot)

# HIV Prevalence and Incidence Rates by Country

inc_country = incidence |>
  group_by(country) |>
  summarize(median = median(inc_rate)) |>
  mutate(country = fct_reorder(country, median)) |>
  plot_ly(x = ~country, y = ~median, type = 'scatter', mode = 'markers') |> 
  layout(title = 'HIV Incidence Rates by Country',
         yaxis = list(title = 'Incidence Rate(%)'))
inc_country


prev_country = prevalence |>
  group_by(country) |>
  summarize(median = median(prev_rate)) |>
  mutate(country = fct_reorder(country, median)) |>
  plot_ly(x = ~country, y = ~median, type = 'scatter', mode = 'markers') |> 
  layout(title = 'HIV Prevalence Rates by Country',
         yaxis = list(title = 'Prevalence Rate(%)'))
prev_country


# HIV Prevalence and Incidence Rates of Each Subpopulation Over Time

year_subpop_inc_plot = incidence |>
  group_by(subpop_pooled,year) |>
  summarise(median = median(inc_rate)) |>
  plot_ly(x = ~year, y = ~median, color = ~subpop_pooled, type = 'scatter', mode = 'lines') |>
    layout(title = 'HIV Incidence Rates of Each Subpopulation Over Year')

print(year_subpop_inc_plot)



```

## Tables
```{r Tables}
# Load necessary libraries
library(dplyr)
library(knitr)
# Top 10 countries with the highest incidence

top_inc = incidence |>
  group_by(country) |>
  summarize(median = median(inc_rate)) |>
  arrange(desc(median)) |>
  top_n(10)


# Top 10 countries with the highest prevalence

top_prev = prevalence |>
  group_by(country) |>
  summarize(median = median(prev_rate)) |>
  arrange(desc(median)) |>
  top_n(10)



# Display the tables with kable for a professional look

kable(top_inc, caption = "Top 10 Countries with Highest HIV Incidence Rates", 
      format = "html", table.attr = "style='width:100%;'", 
      col.names = c("Country", "Median Incidence Rate"))

kable(top_prev, caption = "Top 10 Countries with Highest HIV Prevalence Rates", 
      format = "html", table.attr = "style='width:100%;'", 
      col.names = c("Country", "Median Prevalence Rate"))

```




