---
title: "Sociodemographic Associated with HIV Infection from 1999 to 2016"
output:
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE, message = FALSE}
library(haven)
library(tidyverse)
library(purrr)
library(plotly)
library(modelr)
```

# Data Sources
The [National Health and Nutrition Examination Survey](https://wwwn.cdc.gov/nchs/nhanes/) is designed to evaluate the health and nutritional status of both adults and children. In 1999, the NHANES study transitioned into a continuous format, enabling investigators to conduct longitudinal studies. NHANES encompasses both interview data and physical examination data, which involved information from different perspectives and fields, providing researchers with a comprehensive dataset to examine the prevalence and risk factors associated with various diseases. In this study, we used the demographic data, HIV testing, and sexual behavior data from 1999 to 2016 to identify the potential sociodemographic factors associated with HIV infection status.

# Study Objectives
Our study's primary objective is to analyze the NHANES dataset from 1999 to 2016 to identify sociodemographic factors associated with HIV infection rates in the United States. By examining variables such as age, gender, race, education level, and marial status, we aim to illuminate the complex interplay of societal elements influencing HIV prevalence. This analysis will provide valuable insights into the demographics most affected by HIV, guiding public health strategies and resource allocation to improve prevention and treatment programs. Through this research, we seek to contribute to a more nuanced understanding of HIV epidemiology that could underpin targeted interventions and support the reduction of infection disparities across diverse population groups.


# Data Cleaning

## import NHANES datafiles
Imported demographic, HIV test, and sexual behavior dataframes 1999 - 2016. 
```{r}
demo_list = 
  list.files(path = "NHANES/DEMO", full.names = TRUE) 
hiv_list = 
  list.files(path = "NHANES/HIV", full.names = TRUE) 
sxq_list = 
  list.files(path = "NHANES/SXQ", full.names = TRUE) 


load_files = function(x){
  
  filetype =
    str_extract(x, "DEMO|HIV|SXQ")
  
  year = 
    str_extract(x, "(?<=_)[A-Z]")
  
data = 
    read_xpt(x) |>
    janitor::clean_names()|>
    mutate(filetype = filetype, year = year)
    
}


demo_list_output = 
  map(demo_list, load_files) 

hiv_list_output = 
  map(hiv_list, load_files) 

sxq_list_output = 
  map(sxq_list, load_files)

join_pair = function(df1, df2) {
  left_join(df1, df2, by = c("seqn", "year"))
}


hiv_demo_list = 
  map2(hiv_list_output, demo_list_output, ~join_pair(.x, .y))
  
hiv_demo_sxq_df = 
  map2(hiv_demo_list, sxq_list_output, ~join_pair(.x, .y)) |>
  bind_rows() 

```

Re-coded variables of interest:
Restricted particpants's age between 20 and 49
```{r}
cleaned_joined_df = 
  hiv_demo_sxq_df |>
  filter(ridageyr >= 20 & ridageyr <= 49) |>
  mutate(hiv = case_when(
    lbdhi == 1 | lbxhivc == 1 ~ 1,
    lbdhi == 2 | lbxhivc == 2 ~ 2)
    )|> 
  mutate(MSM = case_when(
    year %in% c("A","B","C") ~ sxq220,
    year %in% c("D","E","F","G","H","I") ~ sxq550)
    )|>
   mutate(WSW = case_when(
    year %in% c("A","B","C") ~ sxq150,
    year %in% c("D","E","F","G","H","I") ~ sxq490)
    )|>
  mutate(year = recode(year, 
                       "A" = "1999-2000",
                       "B" = "2001-2002",
                       "C" = "2003-2004",
                       "D" = "2005-2006",
                       "E" = "2007-2008",
                       "F" = "2009-2010",
                       "G" = "2011-2012",
                       "H" = "2013-2014",
                       "I" = "2015-2016"),
         hiv = recode(hiv,
                            "1" = "Reactive",
                            "2" = "Non-reactive"),
         gender = recode(riagendr, 
                         "1" = "Male",
                         "2" = "Female"),
         age = ridageyr,
         race = recode(ridreth1,
                       "1" = "Mexican American",
                       "2" = "Other Hispanic",
                       "3" = "Non-Hispanic White",
                       "4" = "Non-Hispanic Black",
                       "5" = "Other Race - Including Multi-Racial"),
         education = recode(dmdeduc2,
                            "1" = "Less than 9th grade",
                            "2" = "9-11th grade",
                            "3" = "High school graduate or equivalent",
                            "4" = "Some college or AA degree",
                            "5" = "College graduate or above",
                            "7" = NA_character_,
                            "9" = NA_character_),
         marriage = recode(dmdmartl,
                           "1" = "Married", 
                           "2" = "Widowed",
                           "3" = "Divorced",
                           "4" = "Separated",
                           "5" = "Never married",
                           "6" = "Living with partner",
                           "77" = NA_character_,
                           "99" = NA_character_)
         ) |>
  mutate(samesexcontact = case_when(
    (MSM %in% c(0, 7777, 77777,9999, 99999) | WSW %in% c(0, 7777, 77777, 9999, 99999)) ~ 0,
    (MSM >= 1 & MSM <= 600) | (WSW >= 1 & WSW <= 600) ~ 1
  )) |>
  select(seqn, hiv, age, gender, race, education, marriage, year, samesexcontact)
```




# Descriptive visualization of the potential risk factors
To visualize the association between the selected potential risk factors and HIV test results, we specifically calculated the percent of participants whose HIV-antibody test showing (lbdhi or lbxhivc) ‘positive’ among all the participants who reported to have a ‘positive’ or ‘negative’ value in the antibody test. We further excluded the participants whose status is ‘missing’ or ‘indeterminate’ in the antibody test in our data visualization section.

Then, we plot each bar chart illustrating the integrated (from 1999 to 2016) percentage of ‘positive’ participants on the y-axis and the risk factor of interest on the x-axis, correlated with a line plot indicating the change of the ‘positive’ participants for each risk factor to assess the association between HIV and each risk factor in a rough way. Besides, the mean age of the positive clusters among each sub-cohort was also calculated for each characteristic/risk factor.

## Write a function to combine bar and line plot
We firstly wrote a function for creating the combined plot for each risk factor.
```{r}
combine_plot = function(bar, line, indicator){
  title_g = paste("The Total Percent of HIV-Antibody Positive Patients from 1999 to 2016 by", indicator)
  
  combine_bar = 
    subplot(bar, line, margin = 0.1) %>% 
    layout(title = title_g)
  
  annotations = list( 
  list( 
    x = 0.17,  
    y = 0.98,  
    text = "Total Percent",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),  
  list( 
    x = 0.86,  
    y = 0.98,  
    text = "Percent by Year",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  )) 
  
  combine_bar = combine_bar %>% 
  layout(annotations = annotations)
  
}

```

## Race and the Percent of 'Positive' in HIV Antibody Test 
```{r}
bar_race = 
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(race) %>%
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(race = fct_reorder(race, hiv_percent)) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~race, y = ~hiv_percent, color = ~race, type = "bar", text = ~text_label, colors = "viridis")

line_race =
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(year, race) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(race = fct_reorder(race, hiv_percent)) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~year, y = ~hiv_percent, color = ~race, type = "scatter", mode = "lines+markers",
          text = ~text_label, colors = "viridis") 


combine_race = combine_plot(bar_race, line_race, "Race")

combine_race 
```
The bar plot shown the percent of patients with a positive result from the HIV antibody test in each race when we integrated all datasets from 1999 to 2016. It indicated that the race Non-Hispanic Black has reported the highest percent, while the 'Other Race' had the lowest percent. From the bar plot, disparity among the races.
The line plot shows the change in the reported, positive percent of HIV-antibody test from 1999 to 2016. We could observe that the trend in each race are different, but the Non-Hispanic Black and Other Hispanic had a decreasing trend over 1999 to 2016.
From the two plots, we found that the mean age in each race and each year were also different.



## Sex and the Percent of 'Positive' in HIV Antibody Test 
```{r}
bar_gender = 
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(gender) %>%
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~gender, y = ~hiv_percent, color = ~gender, type = "bar", text = ~text_label, colors = "viridis")


line_gender = 
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(year, gender) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~year, y = ~hiv_percent, color = ~gender, type = "scatter", mode = "lines+markers",
          text = ~text_label, colors = "viridis")

combine_gender = combine_plot(bar_gender, line_gender, "Gender")
combine_gender
```
The bar plot and the line plot indicated that male has reported much higher percent than females. From the line plot, we observed that the positive percent were decreasing in each sex. 
From the two plots, we found that the mean age in females appeared to be slightly higher than males among the patients with a positive test result.



## Education Level and the Percent of 'Positive' in HIV Antibody Test 
```{r}
bar_edu = 
cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(education) %>%
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(education = forcats::fct_relevel(education, c("Less than 9th grade", "9-11th grade", 
                                         "High school graduate or equivalent", 
                                         "Some college or AA degree",
                                         "College graduate or above"))) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~education, y = ~hiv_percent, color = ~education, type = "bar", text = ~text_label, colors = "viridis")


line_edu =
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(year, education) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(education = forcats::fct_relevel(education, c("Less than 9th grade", "9-11th grade", 
                                         "High school graduate or equivalent", 
                                         "Some college or AA degree",
                                         "College graduate or above"))) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~year, y = ~hiv_percent, color = ~education, type = "scatter", mode = "lines+markers",
          text = ~text_label, colors = "viridis") 


combine_edu = combine_plot(bar_edu, line_edu, "Education")
combine_edu 
```
The bar plot indicated that the group with education level of '9-11th grade' has reported the highest percent, while the 'College graduate or above' had the lowest percent; disparity among the education was also observed.
From the line plot, we observed that the trend over 1999 to 2016 in each education level are different, but with a decreasing overall trend. 
From the two plots, we found that the mean age in each race and each year were also different, and it is hard to find a trend.


## Marital status and the Percent of 'Positive' in HIV Antibody Test 
```{r}
bar_mar = 
  cleaned_joined_df %>% 
  drop_na(hiv) %>% 
  group_by(marriage) %>%
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(marriage = fct_reorder(marriage, hiv_percent)) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~marriage, y = ~hiv_percent, color = ~marriage, type = "bar", text = ~text_label, colors = "viridis") 


line_mar =
  cleaned_joined_df %>% 
  drop_na(hiv, marriage) %>% 
  group_by(year, marriage) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(marriage = fct_reorder(marriage, hiv_percent)) %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~year, y = ~hiv_percent, color = ~marriage, type = "scatter", mode = "lines+markers",
          text = ~text_label, colors = "viridis") 

combine_mar = combine_plot(bar_mar, line_mar, "Marital Status")
combine_mar
```
The bar plot indicated that the 'widowed' has reported the highest percent, while the 'married' had the lowest percent.
From the line plot, we observed that the trend over 1999 to 2016 in each education level are different, but with a decreasing overall trend (except for 'widowed', which seemed to be an outlier). 
From the two plots, we found that the mean age in each marital status group and each year also had disparities, and it was hard to find a trend.



## Same-sex Sexual Behavior and the Percent of 'Positive' in HIV Antibody Test 
```{r}
bar_sexbehave = 
  cleaned_joined_df %>%
  drop_na(hiv) %>% 
  mutate(samesexcontact = recode(samesexcontact, "0" = "No same-sex sexual behavior", 
                                  "1" = "Has same-sex sexual behavior")) %>% 
  group_by(samesexcontact) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>% 
  ungroup() %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~samesexcontact, y = ~hiv_percent, color = ~samesexcontact, type = "bar", text = ~text_label, colors = "viridis") 

line_sexbehave =
  cleaned_joined_df %>% 
  drop_na(hiv, samesexcontact) %>% 
  mutate(samesexcontact = recode(samesexcontact, "0" = "No same-sex sexual behavior", 
                                  "1" = "Has same-sex sexual behavior")) %>% 
  group_by(year, samesexcontact) %>% 
  summarize(total_par = n(), 
            age_mean = round(mean(age[hiv == "Reactive"]), 2), 
            hiv_percent = round(sum(hiv == "Reactive") / total_par * 100, 3)) %>%
  ungroup() %>% 
  mutate(text_label = str_c("mean age: ", age_mean, ", Positive HIV Percent: ", hiv_percent, "%")) %>% 
  plot_ly(x = ~year, y = ~hiv_percent, color = ~samesexcontact, type = "scatter", mode = "lines+markers",
          text = ~text_label, colors = "viridis") 


combine_sexbehave = combine_plot(bar_sexbehave, line_sexbehave, "Sexual Behavior")
combine_sexbehave
```
The bar plot and the line plot indicated that the population has same-sex sexual behavior has reported much higher percent than the population without same-sex sexual behavior. From the line plot, we observed that the positive percent fluctuated a lot among the group that has same-sex sexual behavior; while in the group that without same-sex sexual behavior, the the positive percent in each year did not fluctuate as much.
From the two plots, we found that the mean age in the group that has same-sex sexual behavior appeared to be higher than the group without same-sex sexual behavior.


## logistic regression

### Set the reference group
```{r}
cleaned_regression_df= 
   cleaned_joined_df |> 
   mutate(
      hiv_outcome = ifelse(hiv == "Reactive", 1, ifelse(hiv == "Non-reactive", 0, NA)),
      education = forcats::fct_relevel(education, c("Less than 9th grade", "9-11th grade", "High school graduate or equivalent", "Some college or AA degree", "College graduate or above")),
      race = forcats::fct_relevel(race, c("Non-Hispanic White", "Mexican American", "Non-Hispanic Black", "Other Hispanic", "Other Race - Including Multi-Racial")),
      marriage = forcats::fct_relevel(marriage, c("Married", "Widowed", "Divorced", "Separated", "Never married", "Living with partner"))
         )
```

### Bivariate analysis
```{r}
variables = c("age","gender","education","race","samesexcontact","marriage")

fit_and_summarize <- function(var) {
  model = glm(as.formula(paste("hiv_outcome~", var)), data = cleaned_regression_df, family = binomial()) |>
  broom::tidy()
}

model_summaries = 
  map(variables, fit_and_summarize) |>
  bind_rows() |>
  filter(p.value <= 0.008 & term != "(Intercept)")

model_summaries |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 3)
```
Because I am conducting 6 logistic tests, the Bonferroni-corrected significance level would be 0.05/6 = 0.008. Variables `age`, `gender`, `race`, `samesexcontact`, `marriage` were found significant associated with HIV infection status, at 0.8% level of significance. Therefore, these variables were included to the final model.


### The final model
```{r}
fit_regression = cleaned_regression_df |> 
  glm(hiv_outcome ~ samesexcontact + gender + age + race + marriage, data = _, family = binomial()) 

fit_regression|> 
  broom::tidy() |> 
  mutate(OR = exp(estimate)) |> 
  select(term, estimate, OR, p.value)|> 
  knitr::kable(digits = 3)
```
**Logistic regression results:**

* The logistic regression analysis reveals several key risk factors associated with HIV in the US. 
  * Individuals reporting same-sex contacts exhibit a significant increase in the odds of being HIV positive (OR = 17.020), emphasizing the importance of considering sexual behavior in HIV risk assessment. 
  * Gender is also a significant predictor, with males having 6.171 times higher odds of being HIV positive compared to females. 
  * Age demonstrates a positive association, indicating a slight increase in HIV odds with 1-year increase in age (OR = 1.058). 
  * Among racial categories, Non-Hispanic Black individuals show the highest risk (OR = 7.941), while Mexican Americans do not exhibit a statistically significant association. 
  * Regarding marital status, divorced, separated, and never married individuals display elevated odds of being HIV positive, suggesting that certain relationship statuses may contribute to increased vulnerability.
* These findings underscore the complex interplay of demographic factors in HIV risk and contribute valuable insights for targeted prevention and intervention strategies.


### Bootstrap

```{r warning = FALSE}
bootstrap_df =
  cleaned_regression_df |> 
  bootstrap(n = 500) |> 
  mutate(
    models = map(.x = strap, ~glm(hiv_outcome ~ samesexcontact + gender + age + race + marriage, data = .x, family = binomial()) ),
    results = map(models, broom::tidy)) |> 
  select(-strap, -models) |> 
  unnest(results) |> 
  group_by(term) |> 
  mutate(OR = exp(estimate))
```


### Distribution of estimates under repeated sampling
```{r}
bootstrap_df|> 
  filter(term == "age") |> 
  ggplot(aes(x = estimate))+
  geom_density() +
  facet_wrap(.~ term)  
    
bootstrap_df|> 
  filter(term != "age") |> 
  ggplot(aes(x = estimate))+
  geom_density() +
  facet_wrap(.~ term)
```

* This is the uncertainty that we would get in this dataset if we do repeated sampling


### Bootstrap mean OR and 95% CI 
```{r}
bootstrap_df|> 
  summarize(
    mean_OR = mean(OR),
    CI_lower = quantile(OR, 0.025),
    CI_upper = quantile(OR, 0.975)
    )|> 
  knitr::kable(digits = 3)
```
**Bootstrap Confidence Intervals:**

* The bootstrap analysis yields informative insights into the association between various sociodemographic factors and HIV outcomes. 
  * Notably, the wide 95% confidence intervals (CI) for most variables, excluding age, suggest substantial uncertainty in the estimated odds ratios. The variable "samesexcontact" stands out with a mean odds ratio of 18.499 and a relatively narrow CI (10.862 to 31.096), indicating a robust positive association with HIV risk. 
  * In contrast, variables like "genderMale," "marriageDivorced," "marriageLiving with partner," "marriageNever married," "marriageSeparated," "marriageWidowed," "raceMexican American," "raceNon-Hispanic Black," and "raceOther Hispanic" exhibit wide CIs, signifying considerable variability in their associations with HIV. 
* The broad 95% CI underscores the uncertainty in the magnitude of these associations and suggests a need for caution in drawing definitive conclusions. This may be attributed to sample variability or other unaccounted factors influencing the observed relationships. 
  
## Conclusion: 
The logistic regression and bootstrap analyses indicate that several demographic factors are associated with HIV outcomes, same-sex behavior showing a particularly strong positive association. However, the wide 95% confidence intervals for most variables, excluding age, show significant uncertainty in the estimated odds ratios, highlighting the limitations of the study. These broad intervals may stem from sample variability or unaccounted confounders, necessitating caution in drawing definitive conclusions.

Therefore, future research should explore larger and more diverse datasets to enhance the generalizability of findings and address potential sources of variability. Additionally, a thorough investigation into the complex interplay between demographic variables and HIV risk, considering potential interactions and subgroup analyses, could provide a more nuanced understanding of these associations. Moreover, the study could benefit from incorporating behavioral and contextual factors that might contribute to HIV risk.